Mamba-Based Tumor Segmentation\\ \thanks{$\dagger$ Equal contributions}

### 2024-04-24 paper/main.tex
start

\section{Mamba}
Mamba \cite{gu2023mamba} is a recently proposed architecture that combines the modeling power of Transformers with the linear scaling efficiency of state space models (SSMs). The key idea is to augment SSMs with a selection mechanism that allows the model parameters to vary based on the input, enabling it to perform content-aware reasoning.
Standard SSMs map an input sequence $x(t)$ to an output $y(t)$ via a latent state $h(t)$ using a linear time-invariant system:
\begin{align*}
    h'(t) & = \mathbf{A}h(t) + \mathbf{B}x(t) \\
    y(t)  & = \mathbf{C}h(t) + \mathbf{D}x(t)
\end{align*}
\gr{Do these need to be numbered equations? -- I don't. Nvm, disregard unless you think so}

@@

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{DBLP:journals/corr/HeZRS15}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegViT}
\gr{We used SegFormer?}
SegViT \cite{zhang2022segvit} explores the capability of plain Vision Transformers (ViTs) \cite{dosovitskiy2021image} for semantic segmentation, leveraging the attention mechanism to generate masks for semantic segmentation. Unlike previous ViT-based segmentation networks that learn pixel-level representations from the output of the ViT, SegViT introduces the Attention-to-Mask (ATM) module. In the ATM module, the similarity maps between a set of learnable class tokens and the spatial feature maps are transferred to the segmentation masks, enabling effective semantic segmentation. Moreover, to address the computational cost associated with the ViT backbone, SegViT proposes query-based down-sampling (QD) and query-based up-sampling (QU) to build a Shrunk structure. By comparing U-Mamba against SegViT, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the attention-based segmentation mechanism employed by SegViT. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of attention mechanisms and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

@@

\subsection{PP-UMamba}

\gr{To address the suggestion from Tong for "an overview of the whole pipeline" we should probably explain how PP is incorporated. Either by describing the pipeline or making a graphic for the pipeline.}

@@

In our PP-UMamba model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.

% Unsure why this comment line is giving me troubles !!
\gr{$These should all be UL-VM-UNet, or ULVM_UNet if you are against the extra dash. UltraLight VisionMamba-UNet This in each subsection title and in the paragraphs.$}

\subsection{ULMUNet\_v1}
For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder. Additionally, the depth of the U-Net was reduced by one layer to keep the parameter count low while increasing the number of parameters in each step. The intention was to improve performance by increasing the capacity of the model while maintaining a reasonable parameter count.

\subsection{ULMUNet\_v2}
In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8. This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{ultralightvmunet} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.

\subsection{ULMUNet\_v3}
ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2. It increases the number of parallel branches in the PVM block to 8 and modifies the channel list to [16, 32, 64, 128, 256]. Additionally, the depth of the U-Net was reduced to 5 layers. This approach aims to balance the trade-off between model capacity and parameter efficiency.

\subsection{ULMUNet\_v4}
\gr{Pyramidal pooling was used in each of the PVM blocks in the encoder and decoder networks. If we want to do similar to PP-UMamba we can do it just in the bottleneck. Let me know if we want this can I can do it fast.}
This model incorporates pyramidal pooling, similar to PP-UMamba, but within the UltraLight VM-UNet architecture. The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.

\subsection{UL-PP}
\gr{What is UL-PP? v5 is v4 with no PP. If any, v4 is UL-PP.}
The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters. Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.

\section{Results}
We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck \gr{UMamba bottleneck rather than "with"}, UNet, and UltraLight VM-UNet. We train for 100 epochs on all models, and use dice-loss with Adam optimizer. We train on roughly 3000 images, and validate on roughly 400. we then test on roughly 400. We show the stats of our data in Figure \ref{fig:datastats}.
The results are shown in Figures \ref{fig:bars},\ref{fig:balls},\ref{fig:segs}.

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/bars.png}
    \caption{\gr{Thoughts on this vs the two bar graph?}Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
    \label{fig:bars}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/masks.png}
    \caption{Example segmentations.}
    \label{fig:segs}
\end{figure}
\gr{let's include more segmentation results. Maybe in appendix.}

@@

Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy. The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) \gr{how does this outperform? Dice score is lower.}with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.

While the traditional U-Net model achieved the highest dice score of 0.8929, our PP-UMamba model closely followed with a dice score of 0.8867 \gr{note the raise of 300k params in pp?}. Notably, our Mamba-based PP-UMambaBot \gr{name the section in OUR MODELS PP-UMambaBot or change this one. Consistency.} and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.

@@

Future work could involve further refinements and optimizations of the Mamba-based models, as well as their application to other medical imaging modalities and segmentation tasks. Additionally, investigating the interpretability and robustness of these models could provide valuable insights for their deployment in real-world clinical settings.

@@

    \bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
    \bibitem{ultralightvmunet} R. Wu, Y. Liu, P. Liang, and Q. Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr. 20, 2024. [Online]. Available: https://arxiv.org/pdf/2403.20035.pdf
    \bibitem{zhao2017pyramid}
    H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
    \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
\end{thebibliography}
\end{document}

>>>>

\section{Mamba}
Mamba \cite{gu2023mamba} is a recently proposed architecture that combines the modeling power of Transformers with the linear scaling efficiency of state space models (SSMs). The key idea is to augment SSMs with a selection mechanism that allows the model parameters to vary based on the input, enabling it to perform content-aware reasoning.
Standard SSMs map an input sequence $x(t)$ to an output $y(t)$ via a latent state $h(t)$ using a linear time-invariant system:
\begin{align*}
    h'(t) & = \mathbf{A}h(t) + \mathbf{B}x(t) \\
    y(t)  & = \mathbf{C}h(t) + \mathbf{D}x(t)
\end{align*}

@@

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{DBLP:journals/corr/HeZRS15}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegFormer}
SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA). The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks. The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution. This allows SegFormer to efficiently process high-resolution images while capturing both local and global context. Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details. By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

@@

\subsection{UMambaBot\_PPBot}

@@

In our UMambaBot\_PP model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.

% Unsure why this comment line is giving me troubles !!

\subsection{UL-VM-UNet\_v1}
For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder. Additionally, the depth of the U-Net was reduced by one layer to keep the parameter count low while increasing the number of parameters in each step. The intention was to improve performance by increasing the capacity of the model while maintaining a reasonable parameter count.

\subsection{UL-VM-UNet\_v2}
In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8. This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{ultralightvmunet} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.

\subsection{UL-VM-UNet\_v3}
ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2. It increases the number of parallel branches in the PVM block to 8 and modifies the channel list to [16, 32, 64, 128, 256]. Additionally, the depth of the U-Net was reduced to 5 layers. This approach aims to balance the trade-off between model capacity and parameter efficiency.

\subsection{UL-VM-UNet\_v4}
\gr{Pyramidal pooling was used in each of the PVM blocks in the encoder and decoder networks. If we want to do similar to UMambaBot\_PP we can do it just in the bottleneck. Let me know if we want this can I can do it fast.}
This model incorporates pyramidal pooling, similar to UMambaBot\_PP, but within the UltraLight VM-UNet architecture. The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.

\subsection{UL-VM-UNet\_v5}
\gr{What is UL-PP? v5 is v4 with no PP. If any, v4 is UL-PP.}
The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters. Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.

\section{Results}
We now compare our models against ResNet18, ResNet50, standard U-Mamba bottleneck, UNet, and UltraLight VM-UNet. We train for 100 epochs on all models, and use dice-loss with Adam optimizer. We train on roughly 3000 images, and validate on roughly 400. we then test on roughly 400. We show the stats of our data in Figure \ref{fig:datastats}.
The results are shown in Figures \ref{fig:bars},\ref{fig:balls},\ref{fig:segs}.

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|c|c|}
        \hline
        \textbf{Model Name} & \textbf{Parameter Count (Millions)} & \textbf{Dice Score} \\
        \hline
        UMambaBot\_PP & 10.0 & 0.8867 \\
        UMambaBot & 9.8 & 0.8799 \\
        ResNet18 & 13.9 & 0.8648 \\
        ResNet50 & 42.9 & 0.8672 \\
        SegFormer & 17.8 & 0.8454 \\
        UNet & 7.8 & 0.8929 \\
        UL-VM-UNet & 0.049 & 0.79955 \\
        UL-VM-UNet\_v1 & 0.20 & 0.8409 \\
        UL-VM-UNet\_v2 & 0.042 & 0.8243 \\
        UL-VM-UNet\_v3 & 0.176 & 0.8109 \\
        UL-VM-UNet\_v4 & 1.1 & 0.8548 \\
        UL-VM-UNet\_v5 & 0.7 & 0.8633 \\
        \hline
    \end{tabular}
    \caption{Comparison of different segmentation models.}
    \label{tab:model_comparison}
\end{table}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/masks.png}
    \caption{Example segmentations.}
    \label{fig:segs}
\end{figure}

@@

Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy. The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, achieving comparable performance to the larger ResNet50 model (42.9M params) with a dice score of 0.8672, while also outperforming the original UltraLight VM-UNet in terms of both parameter count and dice score.

While the traditional U-Net model achieved the highest dice score of 0.8929, our UMambaBot\_PP model closely followed with a dice score of 0.8867 \gr{note the raise of 300k params in pp?}. Notably, our Mamba-based UMambaBot\_PPBot and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.

@@

Future work could involve further refinements and optimizations of the Mamba-based models, as well as their application to other medical imaging modalities and segmentation tasks. Additionally, investigating the interpretability and robustness of these models could provide valuable insights for their deployment in real-world clinical settings.
\appendix
% Redefine the section numbering for the appendix (optional)
\renewcommand{\thesection}{\Alph{section}}
% First appendix section
\subsection{Training Pipeline}
The training pipeline for our model consists of the following steps:

1. Data preprocessing: The dataset is split into training, validation, and test sets. The images are resized to a uniform size and normalized.

2. Model architecture: We employ a convolutional neural network (CNN) with the following layers: [Describe the layers and their configurations].

3. Training: The model is trained using the training set for [X] epochs with a batch size of [Y]. We use the [Z] optimizer with a learning rate of [L].

4. Evaluation: The trained model is evaluated on the validation set to monitor its performance and prevent overfitting. We use metrics such as accuracy, precision, recall, and F1-score.

5. Testing: Once the model achieves satisfactory performance on the validation set, it is tested on the held-out test set to assess its generalization ability.

% Second appendix section
\subsection{Supplementary Figures}

@@

    \bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
    \bibitem{ultralightvmunet} R. Wu, Y. Liu, P. Liang, and Q. Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr. 20, 2024. [Online]. Available: https://arxiv.org/pdf/2403.20035.pdf
    \bibitem{zhao2017pyramid}
    H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
    \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
    \bibitem{xie2021segformer} E. Xie, W. Wang, Z. Yu, A. Anandkumar, J. Alvarez, and P. Luo, “SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers.” Available: https://arxiv.org/pdf/2105.15203.pdf
\end{thebibliography}
\end{document}
